Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-13 docs/ms-draft.Rmd
updated figs 2 and 5 dimensions, and README (docs/* file uptaded too but I may drop this one off as now the ms lives in ms_preprint/)

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were **standardised** before modelling.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \@ref(modellingdatasets)), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($k$) that describes the data, (ii) re-fit the model by fixing $k$ and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

```{r fig4-partial-plots, fig.cap = "**FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (**Fig. S5**). Refer to Table 1 for the environmental data acronyms.", fig.align = "center", echo = FALSE}

@@

```{r fig2-point-pred, fig.cap = "**FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

@@

```{r fig3-prob-maps, fig.cap = "**FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Results for other seasons and models can be found in the Supplementary Material (**Fig. S6**).", fig.align = "center", echo = FALSE}

>>>>

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \@ref(modellingdatasets)), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($K$) that describes the data, (ii) re-fit the model by fixing $K$ and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

```{r fig2-partial-plots, fig.cap = "**FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (**Fig. S5**). Refer to Table 1 for the environmental data acronyms.", fig.align = "center", echo = FALSE}

@@

```{r fig3-point-pred, fig.cap = "**FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

@@

```{r fig4-prob-maps, fig.cap = "**FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Results for other seasons and models can be found in the Supplementary Material (**Fig. S6**).", fig.align = "center", echo = FALSE}
